\documentclass{amsart}
\input{decls}
\title{Categorical Semantics of Dependent Types}
\author{Frank Tsai}
\date{\today}
% \thanks{}
\begin{document}
\maketitle
\tableofcontents

\newcommand{\Ty}{\ensuremath{\mathcal{U}}}
\newcommand{\Tm}{\ensuremath{\widetilde{\mathcal{U}}}}
\newcommand{\El}{\ensuremath{\mathrm{El}}}
\newcommand{\app}{\ensuremath{\mathrm{App}}}
\newcommand{\Block}[1]{\paragraph{\fbox{\textbf{#1}}}}
\newcommand{\ctxt}{\ensuremath{~\mathsf{ctxt}}}
\newcommand{\type}{\ensuremath{~\mathsf{type}}}
\newcommand{\todo}{{\color{red}{TODO}}}

\newcommand{\Fam}{\ensuremath{\mathsf{Fam}}}
\newcommand{\Cop}{\ensuremath{\mathsf{op}}}

\section{Introduction}
\label{sec:introduction}
This is my personal note for \cite{hofmann:ssdts}, \cite{awodey:natmodels}, and \cite{coraglia:context-judgment-deduction}.
The goal of this note is to survey some existing approaches to categorical semantics in the literature, and hopefully explore the differences between and unity in them.

Categorical semantics provides a framework for semantic investigation of dependent types.
A concrete mathematical model can be identified from an abstract one constructed in the categorical framework.

To develop categorical semantics, certain structures are postulated to avoid distracting ourselves from the task at hand.
As a result, certain complex operations, such as substitution, that one encounters when working with a particular presentation of type theory become primitive ones.

\section{Dependent Type Theory}
\label{sec:dependent-type-theory}

The notion of type theory itself is informal.
However, most would probably agree that a dependent type theory is (informally) a formal system dealing with \emph{types} and \emph{terms} in a \emph{context}, and consisting of the following data:
\begin{itemize}
\item Syntax: a collection of glyphs that make up the atoms of the language.
  These are things such as $\Gamma, A$ or $a : A$.
\item Judgments: sentences that bind together atomic data from the syntax.
  These sentences are intended to describe the status of the given atomic data or how they are related.
  For example,
  \begin{mathpar}
    \vdash \Gamma \ctxt \and \Gamma \vdash A \type \and \Gamma \vdash a : A \and \Gamma \vdash A = B \type \and \Gamma \vdash a = b : A
  \end{mathpar}
  can be informally read as ``$\Gamma$ is a context'', ``$A$ is a type in context $\Gamma$'', ``$a$ has type $A$ in context $\Gamma$'', ``$A$ and $B$ are definitionally equal types in context $\Gamma$'', and ``$a$ and $b$ are definitionally equal terms of type $A$ in context $\Gamma$'', respectively.
\item Rules: a mapping from judgments to judgments.
  The intention of rules is to specify when a new judgment can be made given that some other judgments have already been made.
  For example,
  \begin{mathpar}
    \inferrule*[left=(dTy)]
    { \Gamma \vdash a : A \\ \Gamma.A \vdash B \type }
    { \Gamma \vdash B[a] \type }
  \end{mathpar}
\end{itemize}

Contexts play an important role in dependent type theory.
They can be thought of as the content of a \emph{where} clause in English.
For example, ``$\mathrm{Vec}(x)$ is a type where $x$ is a natural number.''
Then the rule (\textsc{dTy}) yields a new type, say $\mathrm{Vec}(7)$, by substituting $7$ for $x$.

Most type theories that we will discuss feature the following rules regarding contexts and variables.
\begin{mathpar}
  \inferrule*[left=(Empty)]
  { }
  { \diamond \ctxt }
  \and
  \inferrule*[left=(Ctx-Extend)]
  { \Gamma \ctxt \\ \Gamma \vdash A \type }
  { \Gamma, x : A \ctxt }
  \and
  \inferrule*[left=(Var)]
  { \Gamma, x : A \ctxt }
  { \Gamma, x : A \vdash x : A }
\end{mathpar}
These rules should be natural to computer scientists especially the last rule, which states that the variable $x$ has type $A$ when it has been declared to have type $A$.

\subsection{Type Formers}
\label{sec:type-formers}

Type theories can be extended with new types and new terms.
The general scheme for adding new types and new terms is to specify:
\begin{itemize}
\item Formation rules: how to form a new type of this kind.
  For example, given a type $B$ in context $\Gamma.A$, we can form a function type $A \to B$.
\item Introduction rules: how to form a new term of that type.
  For example, given a term $b$ of type $B$ in context $\Gamma.A$, we can form an element of type $A \to B$ via $\lambda$-abstraction: $\lambda_{A}B$.
\item Elimination rules: how to use an element of that type.
  For example, given a dependent function $f$ of type $A \to B$ and a term $a$ of type $A$, we can form an element of type $B$ via function application: $f(a)$.
\item Computation rules ($\beta$-rules): how elimination rules act on introduction rules.
  For example, $(\lambda_{A}b)(a)$ is definitionally equal to $b[a]$.
\item Extensionality principle ($\eta$-rules): how introduction rules act on elimination rules.
  For example, a function $f$ is definitionally equal to $\lambda_{A}(f~x)$.
\end{itemize}

Some quintessential examples of type formers are $\Pi$-types, $\Sigma$-types, $\Id$-types, and universes.
\begin{mathpar}
  \inferrule*
  { \Gamma \vdash A \type \\ \Gamma.A \vdash B \type }
  { \Gamma \vdash \prod_{A}B \type }
  \and
  \inferrule*
  { \Gamma \vdash A \type \\ \Gamma.A \vdash B \type }
  { \Gamma \vdash \sum_{A}B \type }
  \and
  \inferrule*
  { \Gamma \vdash a : A \\ \Gamma \vdash b : A }
  { \Gamma \vdash \Id(a, b) \type }
  \and
  \inferrule*
  { }
  { \Gamma \vdash \cU \type }
\end{mathpar}

\subsection{Examples of Type Theories}
\label{sec:examples-of-type-theories}

In this section, we record a number of type theories.

\subsubsection{Martin-L\"{o}f Type Theory}
\label{sec:martin-lof-type-theory}

Martin-L\"{o}f Type Theory is a collective name for dependent type theories that include several type formers mentioned in \cref{sec:type-formers}, but not a universe closed under impredicative quantification.

\subsubsection{The Logical Framework}
\label{sec:logical-framework}

Martin-L\"{o}f's logical framework is a type theory with $\Pi$-types and a universe.
Its intended use is to define theories as extensions of the LF by constants and equations.
The LF can also be used to encode the syntax of other logical systems, e.g., predicate logic and modal logic.

\subsubsection{The Calculus of Construction}
\label{sec:calculus-of-construction}

The calculus of construction (CoC) is a type theory with $\Pi$-types and a universe $\mathsf{Prop}$ closed under impredicative quantification.
Datatypes defined within $\mathsf{Prop}$ lack induction principles.

\section{Substitution}
\label{sec:substitution}
Write $\Gamma \vdash A \type$ for an element $A$ of the set of semantic types under the semantic context $\Gamma$, and $\Gamma \vdash a : A$ for an element $a$ of the set of semantic terms of type $A$.
For any substitution $\sigma : \Delta \to \Gamma$, write $\blank[\sigma]$ for the action of substitution with $\sigma$ on both types and terms.
We impose the following axioms.
\begin{mathpar}
  A[\sigma \circ \xi] = (A[\sigma])[\xi]
  \and
  a[\sigma \circ \xi] = (a[\sigma])[\xi]
  \and
  A[\id] = A
  \and
  a[\id] = a
\end{mathpar}
These equations correspond to the familiar properties of the inductively defined substitution.
\begin{mathpar}
  A[t[u/x]/x] = (A[t/x])[u/x]
  \and
  a[t[u/x]/x] = (a[t/x])[u/x]
  \and
  A[x/x] = A
  \and
  a[x/x] = a
\end{mathpar}

\section{Category with Families}
\label{sec:category-with-families}

\begin{defn}[Category of families of sets]
  The category $\iF$ of families of sets consists of
  \begin{itemize}
  \item Objects: an indexed family of sets $(I, A_{i \in I})$ where for each $i \in I$, $A_{i}$ is a set.
  \item Morphisms: a morphism $f : (I, A_{i \in I}) \to (J, B_{j \in J})$ consists of a function $f : I \to J$, and an $I$-indexed family of functions $f_{i} : A_{i} \to B_{f(i)}$.
  \end{itemize}
  Identities and compositions are defined in an obvious way.
\end{defn}

\begin{defn}[Category with families]
  A category with families consists of
  \begin{itemize}
  \item A base category $\iC$, whose objects are called contexts and whose morphisms are called substitutions.
  \item A terminal object $\diamond \in \iC$, called the empty context.
  \item A functor $T : \iC\op \to \iF$, mapping each context $\Gamma$ to a family of sets $(\Ty(\Gamma), \Tm_{A \in \Ty(\Gamma)})$, where the set $\Ty(\Gamma)$ is the set of types in context $\Gamma$, and for each $A \in \Ty(\Gamma)$ the set $\Tm_{A}$ is the set of terms of type $A$ in context $\Gamma$.
    For each substitution $\sigma : \Delta \to \Gamma$, $T$ induces a morphism $T(\sigma) : (\Ty(\Gamma), \Tm_{A \in \Ty(\Gamma)}) \to (\Ty(\Delta), \Tm_{A \in \Ty(\Delta)})$, which consists of a function $\blank[\sigma] : \Ty(\Gamma) \to \Ty(\Delta)$, and a family of functions $\blank[\sigma]_{A} : \Tm_{A} \to \Tm_{A[\sigma]}$.
    These functions are the substitution with $\sigma$ on types and on terms, respectively.
    We omit the subscript in the substitution on terms whenever possible.
  \item For each $\Gamma \in \iC$ and each $A \in \Ty(\Gamma)$, there is a context extension $\Gamma.A \in \iC$, a morphism $p_{A} : \Gamma.A \to \Gamma$, and a term $q_{A} \in \Tm(\Gamma.A)_{A[p_{A}]}$ with the universal property that for any $\Delta \in \iC$, any $\sigma : \Delta \to \Gamma$, and any $a \in \Tm(\Delta)_{A[\sigma]}$, there is a unique morphism $(\sigma, a) : \Delta \to \Gamma.A$ such that $p_{A} \circ (\sigma, a) = \sigma$ and that $q_{A}[(\sigma, a)] = a$.
    \input{img/img23}
  \end{itemize}
\end{defn}

Any CwF $\iC$ has the structure of dependent type theory described in \S \ref{sec:dependent-type-theory}.
The terminal object $\diamond \in \iC$ models the empty context.
Context extension and its universal property model the context extension rule and the variable rule.
Additionally, substitutions satisfy the required equations.
This follows directly by functoriality.
For $\sigma : \Delta \to \Gamma$ and $\xi : \Xi \to \Delta$, we have $T(\sigma \circ \xi) = T(\xi) \circ T(\sigma)$.
The left-hand side consists of a function $\blank[\sigma \circ \xi] : \Ty(\Gamma) \to \Ty(\Xi)$, which equals $(\blank[\sigma])[\xi]$ by the right-hand side, and a family of functions $\blank[\sigma \circ \xi]_{A} : \Tm(\Gamma) \to \Tm(\Xi)$, which equals $(\blank[\sigma]_{A[\xi]})[\xi]_{A}$ by the right-hand side.
The other two equations, $A[\id] = A$ and $a[\id] = a$, also follow by functoriality.


\subsection{Terms and Sections}
\label{sec:cwf-terms-and-sections}
Let $\Gamma \in \iC$ and $\Gamma \vdash A \type$.
There is a one-to-one correspondence between the set of sections of $p_{A}$ and the set of terms $\Tm(\Gamma)_{A}$.
Let $\Gamma \vdash a : A$, then evidently $\Gamma \vdash a : A[\id]$.
The universal property for context extension then states that there is a unique section $(\id, a) : \Gamma \to \Gamma.A$.
\[
  \begin{array}{r@{~:~}ccc}
    \varphi & \Tm(\Gamma)_{A} & \to & \mathrm{Section}(p_{A})\\
    \varphi & a & \mapsto & (\id, a)
  \end{array}
\]
Conversely, for any section $\sigma$ of $p_{A}$, there is a term
\[
  q_{A}[\sigma] \in \Tm(\Gamma)_{(A[p_{A}])[\sigma]} = \Tm(\Gamma)_{A[p_{A} \circ \sigma]} = \Tm(\Gamma)_{A}
\]
Thus, we define
\[
  \begin{array}{r@{~:~}ccc}
    \psi & \mathrm{Section}(p_{A}) & \to & \Tm(\Gamma)_{A}\\
    \psi & \sigma & \mapsto & q_{A}[\sigma]
  \end{array}
\]
The two functions are mutual inverses by the universal property of context extension.
Thus, the set of terms of type $A$ in context $\Gamma$ can be viewed as the set of sections of the morphism $p_{A}$.
\input{img/img19}

\subsection{Weakening}
\label{sec:cwf-weakening}
Suppose that $\sigma : \Delta \to \Gamma$ and that $\Gamma \vdash A \type$.
Then $\Delta \vdash A[\sigma] \type$.
Thus, there is a term $\Delta.A[\sigma] \vdash q_{A[\sigma]} : A[\sigma \circ p_{A[\sigma]}]$.
The universal property of context extension gives a substitution $\Delta.A[\sigma] \to \Gamma.A$ as indicated in
\input{img/img16}
We call such a substitution a weakening substitution, and denote it as $\sigma_{A}$.
In the special case where $\sigma$ is $\id_{\Gamma}$, the weakening substitution is the identity.
\input{img/img17}
And in the special case where $\sigma$ is $p_{A}$, the weakening substitution adds a new variable of type $A$ to the context.
In dependent type theory, types in a context can mention variables declared prior to it.
The operation $\blank[p_{A}]$ ``up-shifts'' the variables so that the newly added type $A$ mentions the same variables as the already existing type $A$.
\input{img/img18}

\subsection{Supporting Type Formers}
\label{sec:cwf-supporting-type-formers}

\subsubsection{$\Pi$-Types}
\label{sec:cwf-pi-types}

\begin{defn}
  A CwF supports intensional $\Pi$-types if the following data are given
  \begin{itemize}
  \item An operator $\Pi : \Ty(\Gamma) \times \Ty(\Gamma.A) \to \Ty(\Gamma)$ such that for any $\sigma : \Delta \to \Gamma$
    \[
      \left(\prod_{A}B\right)[\sigma] = \prod_{A[\sigma]}B[\sigma_{A}]
    \]
  \item An operator $\lambda_{A,B} : \Tm(\Gamma.A)_{B} \to \Tm(\Gamma)_{\prod_{A}B}$ such that for any $\sigma : \Delta \to \Gamma$
    \[
      (\lambda_{A,B}b)[\sigma] = \lambda_{A[\sigma],B[\sigma_{A}]}b[\sigma_{A}]
    \]
  \item A morphism $\app_{A,B} : \Gamma.A.(\Pi_{A}B) \to \Gamma.A.B$ such that \input{img/img20} and \input{img/img25} commute.
    And for any morphism $\sigma : \Delta \to \Gamma$,
    \[
      \app_{A,B} \circ \sigma_{A_{(\Pi_{A}B)[\sigma_{A}]}} = \sigma_{A_{B}} \circ \app_{A[\sigma],B[\sigma_{A}]}
    \]
  \end{itemize}
\end{defn}
The morphism $\app_{A,B}$ is rather mysterious.
To unravel what it means, observe that a term $\Gamma \vdash t : B[a]$ can be obtained from the variable $\Gamma.A.B \vdash q_{B} : B[p_{B}]$ with a suitable substitution.

In the elimination rule, two terms $\Gamma \vdash a : A$ and $\Gamma \vdash f : \prod_{A}B$ are given.
The idea is to take $t$ as the application of $f$ to $a$.
The universal property of context extension gives a morphism $(a, f) : \Gamma \to \Gamma.A.(\Pi_{A}B)$.
The first equation yields \input{img/img24}
Thus, there is a term $\Gamma \vdash q_{B}[\app_{A,B} \circ (a,f)] : B[a]$.

Next to validate the $\beta$-rule, the equation $\Gamma \vdash q_{B}[\app_{A,B} \circ (a,\lambda_{A,B}b)] = b[a] : B[a]$ must hold for any $\Gamma.A \vdash b : B$.
The second equation states that $\app_{A,B} \circ (a,\lambda_{A,B}b)$ is precisely $(a,b[a])$, and $q_{B}[a,b[a]] = b[a]$ follows from the universal property of context extension.

Finally, the last equation states that function application is stable under substitution.

\begin{defn}
  A CwF supports extensional $\Pi$-types if it supports intensional $\Pi$-types, and whenever $b \in \Tm(\Gamma.A)_{B}$ and $f \in \Tm(\Gamma)_{\Pi_{A}B}$ and $\app_{A,B} \circ f = b$, then $f = \lambda_{A,B}b$.
\end{defn}

\section{Natural Model}
\label{sec:natural-model}
\begin{defn}[Natural models]
  Let $\iC$ be a small category.
  A natural transformation $\tau : \Tm \to \Ty$ between presheaves on $\iC$ is \emph{representable} when all of its fibers are representable objects in the sense that for any $\Gamma \in C$ and $A \in \Ty(\Gamma)$, we have a pullback.
  \input{img/img01}
  A \emph{natural model} of type theory is a representable natural transformation between presheaves.
  $\Tm$ is the presheaf of terms and $\Ty$ is the presheaf of types.
\end{defn}
The Yoneda Lemma identifies $A : \yon \Gamma \to \Ty$ with a type $A \in \Ty(\Gamma)$ and $b : \yon \Delta \to \Tm$ with a term $b \in \Tm(\Delta)$.
The component $\tau_{\Gamma} : \Tm(\Gamma) \to \Ty(\Gamma)$ can be thought of as the typing of terms in context $\Gamma$.
We write $\Gamma \vdash A \type$ for $A \in \Ty(\Gamma)$ and $\Gamma \vdash b : B$ for $a \in \Tm(\Gamma)$, where $B = \tau \circ b$ as indicated in
\input{img/img02}
Henceforth, we omit $\yon$ whenever possible.

For an arbitrary morphism $\sigma$ in $C$, we use the following notations for the action on morphisms of the respective presheaf.
\begin{mathpar}
  \Ty(\sigma) \iff \blank[\sigma]
  \and
  \Tm(\sigma) \iff \blank[\sigma]
\end{mathpar}
As a consequence of the Yoneda Lemma, $A[\sigma] = A \circ \yon\sigma$ and $a[\sigma] = a \circ \yon\sigma$ for any $\Gamma \vdash A \type$ and $\Gamma \vdash a : A$.

\begin{thm}
  Let $\tau : \Tm \to \Ty$ be a natural transformation of presheaves on a small category $\iC$ with a terminal object.
  Then $\tau$ is representable (a natural model of type theory) just when $(\iC, \tau)$ is a category with families.
\end{thm}

\begin{proof}
  Naturality of $\tau$ means that for any substitution $\sigma : \Delta \to \Gamma$, the diagram
  \input{img/img03}
  commutes.
  Thus, we have that
  \begin{mathpar}
    \Gamma \vdash A \type \imp \Delta \vdash A[\sigma] \type
    \and
    \Gamma \vdash a : A \imp \Delta \vdash a[\sigma] : A[\sigma]
  \end{mathpar}
  The former is given by $\Ty(\sigma)$, and the latter is given by $\Tm(\sigma)$.
  Moreover, the latter ``type-checks'' because the diagram commutes, i.e., $A[\sigma] = (\tau_{\Gamma}(a))[\sigma] = \tau_{\Delta}(a[\sigma])$.

  By functoriality, the following equations are satisfied
  \begin{mathpar}
    A[\sigma \circ \xi] = (A[\sigma])[\xi]
    \and
    A[\id] = A
    \and
    a[\sigma \circ \xi] = (a[\sigma])[\xi]
    \and
    a[\id] = a
  \end{mathpar}
  where $\xi$ is an arbitrary substitution $\xi : \Xi \to \Delta$.
  Note that the contravariance of the presheves $\Tm$ and $\Ty$ plays a role in the order of the compositions, which works out nicely with the chosen notations.
  These data provide a canonical functor $T : \iC\op \to \iF$.
  \input{img/img15}
  As usual, $\tau_{\Delta}\inv(A) \subseteq \Tm(\Delta)$ is the fiber over $A$.
  $T(\sigma)$ consists of the function $\blank[\sigma] : \Ty(\Delta) \to \Ty(\Gamma)$ and a family of functions $\blank[\sigma]_{A} : \tau_{\Delta}\inv(A) \to \tau_{\Gamma}\inv(A[\sigma])$ defined by restricting $\blank[\sigma] : \Tm(\Delta) \to \Tm(\Gamma)$ to the fiber $\tau_{\Delta}\inv(A)$.

  The final piece of data associated with a CwF is context extension, which is encoded precisely by the representability of $\tau$.
  Given any $\Gamma \vdash A \type$, by Yoneda there is a corresponding map $A : \Gamma \to \Ty$.
  Representability of $\tau$ then means that for any $\Delta \vdash a : A[\sigma]$, we have
  \input{img/img04}
  The notations used in the diagram above is justified because $\yon$ is fully faithful.
  In particular, we have that $q \circ \yon(\sigma, a) = a$, which implies that $q_{\Delta} \circ (\sigma, a)_{*} = a_{\Delta}$.
  The component $a_{\Delta} : C(\Delta, \Delta) \to \Tm(\Delta)$ is a function.
  Thus, we have that $q_{\Delta} \circ (\sigma, a)_{*} (\id_{\Delta}) = a_{\Delta}(\id_{\Delta})$.
  By Yoneda, the left-hand side determines the element $q[(\sigma, a)] \in \Tm(\Delta)$, while the right-hand side determines the element $a \in \Tm(\Delta)$.
  Thus, we can extract the required equations from the pullback diagram.
  \begin{mathpar}
    p \circ (\sigma, a) = \sigma
    \and
    q[(\sigma, a)] = a
  \end{mathpar}
  Additionally, by uniqueness, we also have the following required equations.
  \begin{mathpar}
    (\sigma, a) \circ \xi = (\sigma \circ \xi, a[\xi])
    \and
    (p, q) = \id_{\Gamma.A}
  \end{mathpar}
  where $\xi : \Xi \to \Delta$ is an arbitrary substitution.
\end{proof}

By waving a magic wand, every natural model $f : \Tm \to \Ty$ has an associated polynomial functor $F(X) = \sum_{A : \Ty}X^{A}$ such that
\begin{mathpar}
  \mprset{
    fraction={===},
    fractionaboveskip=2.3 ex
  }
  \inferrule
  { (A, B) : \Gamma \to \sum_{A : \Ty}\Ty^{A} }
  { \Gamma \vdash A \type \\ \Gamma.A \vdash B \type }
  \and
  \inferrule
  { (A, b) : \Gamma \to \sum_{A : \Ty}\Tm^{A} }
  { \Gamma \vdash A \type \\ \Gamma.A \vdash b : B }
\end{mathpar}

\subsection{Supporting Type Formers}
\label{sec:nm-supporting-type-formers}

\subsubsection{$\Pi$-Types}
\label{sec:nm-pi-types}

\begin{thm}
  Let $F(X) = \sum_{A : \Ty}X^{A}$ be the polynomial functor associated to a natural model $f : \Tm \to \Ty$.
  Then the type-theoretic rules for (extensional) $\Pi$-types are modeled by maps of the form
  \begin{mathpar}
    \lambda : F(\Tm) \to \Tm
    \and
    \Pi : F(\Ty) \to \Ty
  \end{mathpar}
  making the following diagram a pullback.
  \input{img/img05}
\end{thm}
\begin{proof}
  The type-theoretic $\Pi$-formation rule is
  \begin{mathpar}
    \inferrule
    { \Gamma \vdash A \type \\ \Gamma.A \vdash B \type }
    { \Gamma \vdash \prod_{A}B \type }
  \end{mathpar}
  The premises correspond to a map $(A, B) : \Gamma \to F(\Ty)$.
  Composing $\Pi$ with $(A, B)$ yields $\Gamma \vdash \prod_{A}B \type$.
  \input{img/img06}
  Thus, $\Pi : F(\Ty) \to \Ty$ models the type-theoretic formation rule.

  Similarly, the type-theoretic $\Pi$-introduction rule is
  \begin{mathpar}
    \inferrule
    { \Gamma \vdash A \type \\ \Gamma.A \vdash b : B }
    { \Gamma \vdash \lambda_{A}b : \prod_{A}B }
  \end{mathpar}
  The premises correspond to a map $(A, b) : \Gamma \to F(\Tm)$, whose composition with $\lambda : F(\Tm) \to \Tm$ yields $\Gamma \vdash \lambda_{A}b : \prod_{A}B$.
  Thus, $\lambda : F(\Tm) \to \Tm$ models the type-theoretic introduction rule.
  \input{img/img07}

  Now consider the elimination rule and its associated $\beta$-rule and $\eta$-rule.
  \begin{mathpar}
    \inferrule
    { \Gamma \vdash g : \prod_{A}B \\ \Gamma \vdash a : A }
    { \Gamma \vdash g(a) : B[a] }
    \and
    \inferrule
    { \Gamma.A \vdash b : B \\ \Gamma \vdash a : A }
    { \Gamma \vdash (\lambda_{A}b)(a) = b[a] : B[a] }
    \and
    \inferrule
    { \Gamma \vdash g : \prod_{A}B }
    { \Gamma \vdash \lambda_{A}(g[p_{A}](q_{A})) = g : \prod_{A}B }
  \end{mathpar}
  where $b[a]$ and $B[a]$ are defined respectively as
  \begin{mathpar}
    b \circ (1, a) \and B \circ (1, a)
  \end{mathpar}
  as indicated in
  \input{img/img08}
  Intuitively, $B[a]$ corresponds to the type obtained by filling in $a$ for the last variable in the context, and $b[a]$ corresponds to the term obtained by filling in $a$ for the last variable in the context.
  
  The first premise of the elimination rule states that we have a commutative diagram.
  \input{img/img09}
  By the universal property of the pullback, we then have a unique map $(A, \widetilde{g}) : \Gamma \to F(\Tm)$, corresponding to $\Gamma \vdash A \type$ and $\Gamma.A \vdash \widetilde{g} : B$.
  \input{img/img10}
  Thus, we have a term $\Gamma \vdash \widetilde{g}[a] : B[a]$, and we take $g(a) := \gtil[a]$.
  \input{img/img11}

  From the first premise of the $\beta$-rule, we can extract that
  \begin{mathpar}
    \Gamma \vdash A \type
    \and
    \Gamma.A \vdash B \type
    \and
    \Gamma.A \vdash b : B
  \end{mathpar}
  Thus, there are maps $(A, B) : \Gamma \to F(\Ty)$ and $(A, b) : \Gamma \to F(\Tm)$.
  The introduction rule gives a term $\Gamma \vdash \lambda_{A}b : B$ so that the diagram commutes.
  \input{img/img12}
  Thus, there is a unique map $(A, \widetilde{\lambda_{A}b}) : \Gamma \to F(\Tm)$.
  \input{img/img13}
  By definition, $\Gamma \vdash (\lambda_{A}b)(a) = \widetilde{\lambda_{A}b}[a] : B[a]$, where $a$ is given by the second premise.
  Additionally, $\lambda \circ (A, b) = \lambda_{A}b$.
  By uniqueness, $\Gamma.A \vdash b = \widetilde{\lambda_{A}b} : B$, yielding the desired equation: $\Gamma \vdash (\lambda_{A}b)(a) = b[a] : B[a]$.

  Finally, the $\eta$-law is also satisfied.
  The premise says that there is a unique map $(A, \gtil) \maps \Gamma \to F(\Tm)$.
  \input{img/img14}
  By definition, $(g[p_{A}])(q_{A}) = \widetilde{g[p_{A}]}[q_{A}] = \widetilde{g[p_{A}]} \circ (1, q_{A})$.

  The type and term constructors $\Pi$, $\lambda$, and $\blank(\blank)$ also respect substitutions $\sigma \maps \Delta \to \Gamma$.
  Consider the formation rule,
  \begin{mathpar}
    \inferrule
    { \Gamma \vdash A \type \\ \Gamma.A \vdash B \type }
    { \Gamma \vdash \prod_{A}B \type }
  \end{mathpar}
  Applying $\sigma$ to each premise yields a new instance of the formation rule.

  $\Gamma.A \vdash B \type$,
  $\Delta.A[\sigma] \vdash B[\sigma] \type$
  
  \begin{mathpar}
    \inferrule
    { \Delta \vdash A[\sigma] \type \\ \Delta.A[\sigma] \vdash B[\sigma] \type }
    { \Delta \vdash \prod_{A[\sigma]}B[\sigma] \type }
  \end{mathpar}
\end{proof}

\section{Judgmental Theory}
\label{sec:judgmental-theory}
\begin{defn}
  A \emph{prejudgmental theory} consists of the following data:
  \begin{itemize}
  \item a category $\iC$ with a terminal object $\diamond$;
  \item a set $\cJ$ of functors $f : \dF \to \iC$;
  \item a set $\cR$ of functors $\lambda : \dF \to \dG$;
  \item a set $\cP$ of 2-cells;
  \end{itemize}
\end{defn}

\paragraph{\textbf{Terminology}}
\begin{itemize}
\item The objects of the category $\iC$ are called contexts and the morphisms are substitutions.
  The terminal object is called the empty context.
\item An element $f : \dF \to \iC$ of $\cJ$ is a \emph{classifier for the judgment} $\dF$.
  The boundary between a judgment and a judgment classifier is often blurred.
\item A \emph{rule} is an element $\lambda$ of $\cR$.
\item A \emph{policy} is an element $\lambda^{\#}$ of $\cP$.
\end{itemize}

\begin{defn}
  A \emph{judgmental theory} is a prejudgmental theory such that the following coherence conditions are satisfied:
  \begin{itemize}
  \item $\cR$ and $\cP$ are closed under composition;
  \item the judgments are precisely those rules whose codomain is $\iC$;
  \item $\cR$ and $\cP$ are closed under finite limits, $\#$-liftings, and whiskering.
  \end{itemize}
  A classifier is said to be \emph{(op)substitutional} if it is a Grothendieck (op)fibration.
\end{defn}

\begin{rmk}
  Both the slice and the coslice over an object $\Gamma \in \iC$ can be obtained by pulling back the universal discrete opfibration $\textsf{Set}_{*} \to \textsf{Set}$ along the (co)representable functor.
  Thus, the coslice is a discrete Grothendieck opfibration, while the slice is a discrete Grothendieck fibration.
\end{rmk}

\begin{defn}
  A (pre)judgmental theory has axioms if the initial fibration over $\iC$ is a judgment.
\end{defn}

\begin{defn}
  A (pre)judgmental theory has substitution if
  \begin{itemize}
  \item all slices are judgments;
  \item for all fibrations $\dF \in \cR$ and for all $F \in f\inv(\Gamma)$, the corresponding arrow $F : \iC/\Gamma \to \dF$ is a rule.
  \end{itemize}
\end{defn}

\subsection{Judgmental calculi}
\label{sec:judgmental-calculi}

Recall that a dependent type theory consists of syntax, judgments, and rules.
Given a judgmental theory, the corresponding judgmental calculus provides these data.
\begin{itemize}
\item Syntax: a letter $\Gamma$ for each context in $\iC$, and a letter $F$ for each object in a judgment classifier $\dF$.
\item Judgments: there are two main kinds of judgment for each classifier $\dG \in \cJ$.
  \begin{itemize}
  \item The first kind of judgment acknowledges a piece of $\dG$-empirical evidence and clarifies the status of an object.
    For $g : \dG \to \iC$, we write
    \[
      \Gamma \vdash G~\dG(g)
    \]
    for an object in the fiber category $\dG_{\Gamma}$.
    This can be understood as, for example, ``$G$ is green given $\Gamma$''.
  \item The second kind of judgment is an identity checker.
    For two equal objects $G, G' \in \dG_{\Gamma}$, we write
    \[
      \Gamma \vdash G =_{\dG(g)} G'
    \]
    This can be understood as ``$G$ and $G'$ are indistinguishable by green given $\Gamma$''.
  \end{itemize}
\end{itemize}

\subsection{Nested Judgments}
\label{sec:nested-judgments}

Often, we find ourselves in a situation where we need to classify a pair of judgments that are related in some way.
For example, in the identity formation rule the premise is a pair of judgments, $\Gamma \vdash a : A$ and $\Gamma \vdash b : B$, under the same context.
A discussion of identity types is deferred to a later section. 

Say, we wish to classify the pair of judgments: $\Gamma \vdash H~\dH(g\lambda)$ and $g\lambda(H) \vdash F~\dF(f)$.
Note that $H$ and $F$ form a cone over the cospan $g\lambda$ and $f$.
\input{img/nested01}
Thus, by the universal property of pullbacks there is a one-to-one correspondence
\begin{mathpar}
  \mprset{
    fraction={===}
  }
  \inferrule
  { \Gamma \vdash H.\lambda F~\dH.\lambda\dF }
  { \Gamma \vdash H~\dH \\ g\lambda(H) \vdash F~\dF }
\end{mathpar}
That is, the nested judgment classifier $\dH.\lambda\dF$ classifies a pair of judgments of the form $\Gamma \vdash H~\dH$ and $g\lambda(H) \vdash F~\dF$.

\subsection{Rules}
\label{sec:jt-rules}

Given a rule $\lambda : \dF \to \dG$ and $\Gamma \vdash F~\dF$, then one has $g\lambda(F) \vdash \lambda(F)~\dG$.
\begin{mathpar}
  \input{img/jt-rules01} \and (\lambda)\,\inferrule{ \Gamma \vdash F~\dF }{ g\lambda(F) \vdash \lambda(F)~\dG }
\end{mathpar}

\subsection{Substitutionality}
\label{sec:jt-substitutionality}

In a substitutional judgmental theory, the corresponding functor of an object $\Gamma \vdash F~\dF$ is a rule indicated as follows.
Thus, for any substitution $\sigma : \Delta \to \Gamma$ in $\iC$, one has $\Delta \vdash F(\sigma)~\dF$.
\begin{mathpar}
  \input{img/jt-subst01} \and (F)\,\inferrule{ \Delta \vdash \sigma~\iC/\Gamma }{ \Delta \vdash F(\sigma)~\dF }
\end{mathpar}

\subsection{Policies}
\label{sec:jt-policies}

Suppose that we are given a substitutional judgmental theory with
\input{img/jt-policy01}
Then for any $\Gamma \vdash F~\dF$, there is a substitution $\lambda^{\#}_{F} : g\lambda(F) \to \Gamma$.
Thus,
\begin{mathpar}
  \inferrule
  { \Gamma \vdash F~\dF }
  { g\lambda(F) \vdash F(\lambda^{\#}_{F})~\dF }
\end{mathpar}

\subsection{$\#$-Lifting}
\label{sec:sharp-lifting}

Let $\lambda^{\#}$ be a policy and $\cR^{*}$ be a rule which happens to be a fibration depicted as follows.
\input{img/sharp-lifting01}

We can pull back two maps along $\cR^{*}$.
\input{img/sharp-lifting02}

Then $\lambda$ and $\lambda^{\#}$ can be lifted to complete the triangle above.
\input{img/sharp-lifting04}

\section{Dependent Type Theory in Judgmental Theory}
\label{sec:dependent-type-theory-in-judgmental-theory}

\begin{defn}
  A dependent type theory is a substitutional judgmental theory generated by
  \input{img/jddt01}
\end{defn}

We can think of $\Tm$ and $\Ty$ as the categories classifying terms and types, respectively.
The functor $\Sigma$ performs typing, while the functor $\Delta$ performs context extension.
Indeed, given $\Gamma \vdash a~\Tm$, one has $\Gamma \vdash \Sigma(a)~\Ty$, which is precisely the right-hand rule in a less familiar notation. 
\begin{mathpar}
  \inferrule
  { \Gamma \vdash a~\Tm }
  { \Gamma \vdash \Sigma(a)~\Ty }
  \and
  \inferrule
  { \Gamma \vdash a : A }
  { \Gamma \vdash A \type }
\end{mathpar}

By whiskering the following diagram
\input{img/jddt02}
one obtains the following policy
\input{img/jddt03}
whose components $\delta_{A}$ are given by $u(\epsilon_{A})$.
Thus, given $\Gamma \vdash A~\Ty$, one obtains $\util\Delta(A) \vdash A(\delta_{A})~\Ty$ and $\util\Delta(A) \vdash \Delta(A)~\Tm$.
\begin{mathpar}
  \inferrule
  { \Gamma \vdash A~\Ty }
  { \util\Delta(A) \vdash \Delta(A)~\Tm }
  \and
  \inferrule
  { \Gamma \vdash A \type }
  { \Gamma.A \vdash q_{A} : A(\delta_{A}) }
\end{mathpar}
The following lemma shows that $\Delta(A)$ has the expected type.

\begin{lem}[Emergence of Substitution]
  Let $A \in \Ty$ be an object.
  Then $\Sigma\Delta(A) = A(\delta_{A})$.
  Here, the $A$ on the right-hand side is given by the fibrational Yoneda Lemma.
\end{lem}
\begin{proof}
  The Cartesian lift of $\delta_{A} : u(\Sigma\Delta(A)) \to u(A)$ is a morphism $\widetilde{\delta_{A}} : A(\delta_{A}) \to A$.
  It is then sufficient to show that the Cartesian lift of $\delta_{A}$ is a map $f : \Sigma\Delta(A) \to A$.
  This follows immediately from the construction of $\delta_{A}$ as $\epsilon_{A} : \Sigma\Delta(A) \to A$ is precisely the sought-after morphism.
\end{proof}

The classifier $\Ty.\Delta\Ty$ given by the pullback
\input{img/jddt05}
classifies a pair of judgments of the form $\Gamma \vdash A~\Ty$ and $\util\Delta(A) \vdash B~\Ty$.
The map $u.\util\Delta$ projects out $\Gamma \vdash A~\Ty$, which ends up landing in $\Gamma.A$ after traveling along $\util\Delta$.
Alternatively, one can travel along $u$ directly without visiting the middleman $\Tm$, ending up in $\Gamma$.
These two paths are related by a policy $\delta'$ obtained by whiskering the following diagram.
\input{img/jdtt01}
Henceforth, we let $v = u \circ u.\util\Delta$.

\subsection{Type Dependency}
\label{sec:type-dependency}

In dependent type theory, if one has a type $B$ that depends on $A$ and an element of $A$, then one can obtain a new type $B[a]$ by substituting $a$ into $B$.
\begin{mathpar}
  \inferrule
  { \Gamma \vdash a : A \\ \Gamma.A \vdash B \type }
  { \Gamma \vdash B[a] \type }
  \and
  \inferrule
  { \Gamma \vdash a : A \\ \Gamma.A \vdash b : B }
  { \Gamma \vdash b[a] : B[a] }
\end{mathpar}
First, we need to classify the premises.
\input{img/jddt04}
Then for $\Gamma \vdash a : A$ and $\Gamma.A \vdash B \type$, we have $\Gamma.A \vdash (a, B)~\Tm.\Sigma\Delta\Ty$.

Let $\eta' = \util\eta$.
Then each component $\eta'_{a} : \Gamma \to \Gamma.A$ is a substitution, corresponding to the substitution $\blank[a]$.
\input{img/etaprime01}

This triangle can be lifted to a triangle involving the classifier $\Tm.\Sigma\Delta\Ty$.
\input{img/jddt06}

A pair of judgments $\Gamma \vdash a : A$ and $\Gamma.A \vdash B \type$ corresponds uniquely to an object $(a, B) \in \Tm.\Sigma\Delta\Ty$.
Traversing along $\util\Delta\Sigma.u$ projects out $B \in \Ty$, which lives in the fiber over $\Gamma.A$.
Thus, the object $(a, B)$ as a whole lives in the fiber over $\Gamma.A$, yielding
\begin{mathpar}
  \inferrule
  { \Gamma.A \vdash (a,B)~\Tm.\Sigma\Delta\Ty }
  { \Gamma.A \vdash (a,B)(u^{*}\eta')~\Tm.\Sigma\Delta\Ty(u \circ \util\Delta\Sigma.u) }
\end{mathpar}
We want the object to live in the fiber over $\Gamma$ instead of $\Gamma.A$.
This can be done by adjusting the policy as follows.
\input{img/jddt07}
Thus, one has
\begin{mathpar}
  \inferrule
  { \Gamma.A \vdash (a,B)~\Tm.\Sigma\Delta\Ty }
  { \Gamma \vdash (a,B[a])~\Tm.\Sigma\Delta\Ty(v) }
\end{mathpar}

\subsection{Supporting Type Formers}
\label{sec:jdtt-supporting-type-formers}

\subsection{$\Pi$-Types}
\label{sec:jddt-pi-types}

Having $\Pi$-types means to implement the following rules:
\begin{mathpar}
  \inferrule
  { \Gamma \vdash A \type \\ \Gamma.A \vdash B \type }
  { \Gamma \vdash \Pi_{A}B \type }
  \and
  \inferrule
  { \Gamma \vdash A \type \\ \Gamma.A \vdash b : B }
  { \Gamma \vdash \lambda_{A}b : \Pi_{A}B }
  \and
  \inferrule
  { \Gamma \vdash f : \Pi_{A}B \\ \Gamma \vdash a : A }
  { \Gamma \vdash f(a) : B[a] }
  \and
  \inferrule
  { \Gamma.A \vdash b : B \\ \Gamma \vdash a : A }
  { \Gamma \vdash (\lambda_{A}b)(a) = b[a] : B[a] }
\end{mathpar}

The classifiers for the premises of the formation and the introduction rules are constructed as indicated below.
\input{img/jddt-pi01}

A dependent type theory has $\Pi$-types if the following is a commutative diagram.
\input{img/jddt-pi02}

The formation rule is precisely $\Pi$ and the introduction rule is precisely $\lambda$.
For the other rules, we need to classify
\begin{mathpar}
  \Gamma \vdash A \type \and \Gamma.A \vdash B \type \and \Gamma \vdash a : A \and \Gamma \vdash f : \Pi_{A}B
\end{mathpar}
It is clear that $\Tm.\Sigma\Delta\Ty$ classifies the first three.
The last judgment is classified by pulling back $\Sigma$ along the composite $\Pi' = \Pi \circ \pi$.

\input{img/jddt-pi03}

\bibliographystyle{alpha}
\bibliography{all}

\end{document}

% LocalWords:  prejudgmental
