\documentclass{amsart}
\input{decls}
\title{Canonical Presentations via Free Algebras}
\author{Frank Tsai}
\date{\today}
%\thanks{}
\begin{document}
\maketitle
\tableofcontents

\section{Introduction}
\label{sec:introduction}

\begin{defn}
  A \emph{split coequalizer} is a diagram
  \input{img/4-01}
  so that
  \begin{enumerate}
  \item $(f,g,h)$ defines a \emph{fork}, i.e., $h$ defines a cocone under the parallel morphisms $x \toto y$.
  \item $h$ is a retraction for $s$, i.e., $hs = \id_{z}$.
  \item $g$ is a retraction for $t$, i.e., $gt = \id_{y}$.
  \item $ft = sh$.
  \end{enumerate}
\end{defn}

\begin{lem}\label{lem:absolute-coequalizer}
  Any split coequalizer is an \emph{absolute coequalizer}, i.e., any functor preserves it.
\end{lem}
\begin{proof}
  For any fork $(f,g,u)$, we wish to show that there is a unique way to factorize $u$.
  Since $h$ is a split epimorphism, the uniqueness of any such factorization is immediate.
  It remains to show that a factorization always exists.
  A natural candidate is $u \circ s$.
  \input{img/4-02}
  It remains to show that $ush = u$.
  \begin{align}
    ush &= uft\\
        &= ugt\\
        &= u
  \end{align}
  Since functors preserve commutative diagram, the image of a split coequalizer under any functor is a split coequalizer.
\end{proof}

\begin{defn}
  Given a functor $U : \iD \to \iC$:
  \begin{enumerate}
  \item A \emph{$U$-split coequalizer} is a parallel pair $f,g : x \toto y$ together with an extension of the pair $Uf,Ug : Ux \toto Uy$ to a split coequalizer.
    \input{img/4-03}
    in $\iC$.
  \item $U$ \emph{creates coequalizers of $U$-split pairs} if every $U$-split coequalizer $(f,g,h,s,t)$ in $\iC$ admits a lifting: $f,g$ admit a coequalizer in $\iD$ whose image under $U$ is isomorphic to the fork $(Uf,Ug,h)$.
  \end{enumerate}
\end{defn}

\begin{lem}\label{lem:forgetful-strictly-creates}
  For any monad $(T, \eta, \mu)$ acting on $\iC$, the monadic forgetful functor $U^{T} : \iC^{T} \to \iC$ strictly creates coequalizers of $U^{T}$-split pairs.
\end{lem}
\begin{proof}
  Given any parallel pair $f,g : (A,\alpha) \toto (B,\beta)$ in $\iC^{T}$ that admits a $U^{T}$-splitting:
  \input{img/4-04}
  in $\iC$.
  We need to show that $C$ can be given a unique algebra structure $\gamma$ so that $h : (B,\beta) \to (C,\gamma)$ is a coequalizer.
  A $U^{T}$-splitting is an absolute coequalizer.
  Thus, the image of the $U^{T}$-splitting under $U$ is another absolute coequalizer.
  Since $f$ and $g$ are $T$-homomorphisms, the left-hand square commutes.
  \input{img/4-05}
  Clearly, $(Tf,Tg,h\beta)$ is a fork.
  Thus, there is a unique map $\gamma$ that makes the right-hand square commute.
  If $\gamma$ is an algebra structure, then this demonstrates that $h$ is a $T$-homomorphism, and $\gamma$ is the only algebra structure that makes $h$ a $T$-homomorphism.

  We need to check that the following diagrams commute:
  \begin{mathpar}
    \input{img/4-06} \and \input{img/4-07}
  \end{mathpar}
  
  By \cref{lem:absolute-coequalizer}, $\gamma = h\beta\-Ts$.
  Thus,
  \begin{align}
    \gamma\eta_{C} &= h\beta\-(Ts)\eta_{C}\\
           &= h\beta\eta_{B}s\\
           &= hs\\
           &= \id_{C}\\
    \gamma\mu_{C} &= h\beta(Ts)\mu_{C}\\
           &= h\beta\mu_{B}T^{2}s\\
    \gamma\-T\gamma &= h\beta(Ts)(Th)(T\beta)T^{2}s\\
           &= h\beta(Tf)(Tt)(T\beta)T^{2}s\\
           &= h\beta(Tg)(Tt)(T\beta)T^{2}s\\
           &= h\beta(T\beta)T^{2}s\\
           &= h\beta\mu_{B}T^{2}s
  \end{align}

  It remains to show that $h : (B,\beta) \to (C,\gamma)$ is a coequalizer.
  That is, for any $T$-homomorphism $k : (B,\beta) \to (D,\delta)$ creating a fork $(f,g,h)$, $k$ admits a unique factorization.
  Note that $k$ admits a unique factorization in $\iC$ by the universal property of coequalizer.
  \input{img/4-08}
  Thus, it suffices to show that $j$ is a $T$-homomorphism.
  That is,
  \input{img/4-09}
  \begin{align}
    j\gamma &= jh\beta\-Ts\\
       &= k\beta\-Ts\\
       &= \delta\-TkTs\\
       &= \delta\-Tj
  \end{align}
\end{proof}

\begin{lem}
  Let $(T,\eta,\mu)$ be a monad on $\iC$, and let $(A,\alpha)$ be a $T$-algebra.
  Then
  \input{img/4-10}
  is a coequalizer in $\iC^{T}$.
\end{lem}
\begin{proof}
  By \cref{lem:forgetful-strictly-creates}, it suffices to show that $\alpha : TA \to A$ and $T\alpha,\mu_{A} : T^{2}A \toto TA$ can be extended to a split coequalizer.
  Consider
  \input{img/4-11}
  Clearly, $\eta_{A}$ (resp., $\eta_{TA}$) is a retraction for $\alpha$ (resp., $\mu_{A}$).
  The required equation $\eta_{A} \circ \alpha = T\alpha \circ \eta_{TA}$ is just the naturality of $\eta$.
\end{proof}

\begin{cor}
  If $F,U : \iC \toot \iD$ is a monadic adjunction, then
  \begin{enumerate}
  \item $U : \iD \to \iC$ creates coequalizers of $U$-split pairs;
  \item For any $D \in \iD$, there is a coequalizer
    \input{img/4-12}
  \end{enumerate}
\end{cor}
\begin{proof}
  Since the adjunction is monadic, the canonical comparison functor $K : \iD \to \iC^{T}$ defines an equivalence.
  \input{img/4-13}
  Since $U = U^{T}K$, for any $U$-split pair $f,g : x \toto y$ in $\iD$, $Kf,Kg : Kx \toto Ky$ is a $U^{T}$-split pair.
  By \cref{lem:forgetful-strictly-creates}, $Kf,Kg$ can be extended to a coequalizer in $\iC^{T}$, which transports to a coequalizer in $\iD$ via any inverse equivalence of $K$.

  For (ii), it suffices to show that $FU\epsilon_{D}, \epsilon_{FUD}$ is a $U$-split pair.
  Indeed, they are.
  \input{img/4-14}
  All the equations hold due to the triangle equalities.
\end{proof}

\bibliographystyle{alpha}
\bibliography{all}

\end{document}
