\documentclass{amsart}
\input{decls}
\title{Limits and Colimits in Categories of Algebras}
\author{Frank Tsai}
\date{\today}
%\thanks{}
\begin{document}
\maketitle
\tableofcontents

\section{Introduction}
\label{sec:introduction}

\begin{lem}
  If $U : \iA \to \iC$ is monadic, then $U$ is conservative.
\end{lem}
\begin{proof}
  Since $U$ is monadic, the canonical comparison functor $K : \iA \to \iC^{T}$ is fully faithful, and thus conservative.
  And since $U = U^{T} \circ K$, it suffices to show that $U^{T}$ is conservative.

  Any $T$-homomorphism $f : (A,\alpha) \to (B,\beta)$ is a morphism $f : A \to B$.
  If $f$ is an isomorphism, then there is an inverse $f\inv : B \to A$, and we wish to show that $f\inv$ is a $T$-homomorphism.
  That is,
  \[
    f\inv \circ \beta = \alpha \circ Tf\inv
  \]
  Since $f$ is an isomorphism (so a monomorphism), it suffices to show that
  \[
    f \circ f\inv \beta = f \circ \alpha \circ Tf\inv
  \]
  This follows immediately from the fact that $f$ is a $T$-homomorphism.
\end{proof}

\begin{cor}
  Any bijective homomorphism in a category of models for an algebraic theory is an isomorphism.
\end{cor}

\begin{thm}\label{thm:monadic-create-limits}
  A monadic functor $U : \iA \to \iC$ creates
  \begin{enumerate}
  \item all limits that $\iC$ has, and
  \item all colimits that $\iC$ has and $T$ and $T^{2}$ preserve.
  \end{enumerate}
\end{thm}
\begin{proof}
  Since $\iA \eqv \iC^{T}$, the comparison functor creates any limit and colimit of $\iC^{T}$.
  Thus, it suffices to prove that $U^{T}$ has these properties.
  
  For (i), consider a diagram $D : \iJ \to \iC^{T}$ spanning $(Dj_{i},\gamma_{i}) \in \iC^{T}$, whose image under $U^{T}$ is a limit cone $\lambda : \ell \to U^{T}D$.
  We need to lift $\ell$ to a $T$-algebra.
  Note that we can define a cone over $U^{T}D$ as follows:
  \input{img/6-01}
  This suggests that the algebra map should be the map $\alpha : T\ell \to \ell$ given by the universal property of limits.
  It is completely obvious that $(\ell,\alpha)$ is a limit since for any cone $\nu : (D,\kappa) \to D$, the unique map $\nu' : D \to \ell$ is a $T$-homomorphism.
  It remains to check that $(\ell,\alpha)$ is an algebra.
  
  Consider the following cone:
  \input{img/6-02}
  Clearly, $\id_{\ell}$ factors through the legs of the cone.
  Thus, $\alpha \circ \eta_{\ell} = \id_{\ell}$.
  Now, consider the following:
  \input{img/6-03}
  Since $\alpha \circ \mu_{\ell}$ and $\alpha \circ T\alpha$ factor through the legs of the same cone, uniqueness demands that they are equal.

  For (ii), consider a diagram $D : \iJ \to \iC^{T}$ spanning $(Dj_{i},\gamma_{i}) \in \iC^{T}$, whose image under $U^{T}$ is a colimit $\lambda : U^{T}D \to \ell$ preserved by $T$ and $T^{2}$.
  Our goal is to lift $\ell$ to a $T$-algebra defining a colimit of $D$.
  Consider the following tower of colimits:
  \input{img/6-04}
  Define $\alpha := \zeta \circ \xi$.
  Again, it is clear that $(\ell,\alpha)$ is a colimit if $\alpha$ is an algebra map.
  Thus, it remains to check that $(\ell,\alpha)$ is an algebra.
  Consider
  \input{img/6-05}
  Clearly, $\id_{\ell}$ factors through this cocone.
  Thus, uniqueness demands that $\alpha \circ \eta_{\ell} = \id_{\ell}$.
  Also consider
  \input{img/6-06}
  Uniqueness demands that $\alpha \circ \mu_{\ell} = \alpha \circ T\alpha$.
\end{proof}

\cref{thm:monadic-create-limits} has a number of corollaries.

\begin{cor}
  The inclusion functor of a reflexive subcategory creates all limits.
\end{cor}
\begin{proof}
  The inclusion functor of a reflexive subcategory is monadic.
  Thus, the result follows immediately from \cref{thm:monadic-create-limits}
\end{proof}

\begin{cor}\label{cor:monadic-set-complete}
  Any category that is monadic over $\mathsf{Set}$ is complete.
\end{cor}
\begin{proof}
  This is immediate since $\mathsf{Set}$ is complete.
\end{proof}

\begin{cor}
  $\mathsf{Set}$ is cocomplete.
\end{cor}
\begin{proof}
  We have already proved this.
  An alternate proof of this fact that relies on \cref{thm:monadic-create-limits} is based on the observation that the contravariant powerset functor $\cP : \mathsf{Set}\op \to \mathsf{Set}$ is monadic.
  Thus, by \cref{cor:monadic-set-complete}, $\mathsf{Set}\op$ is complete, proving that $\mathsf{Set}$ is cocomplete.
\end{proof}

\cref{thm:monadic-create-limits} shows that any category of models for an algebraic theory is complete.
In fact, any such a category is also cocomplete.

\begin{thm}
  Suppose $\iC$ is cocomplete and $U : \iA \to \iC$ is monadic.
  Then the following are equivalent:
  \begin{enumerate}
  \item $\iA$ is cocomplete.
  \item $\iA$ has coequalizers.
  \end{enumerate}
\end{thm}

\bibliographystyle{alpha}
\bibliography{all}

\end{document}
