\documentclass{amsart}
\input{decls}
\title{A Formula for Kan Extentions}
\author{Frank Tsai}
\date{\today}
%\thanks{}
\begin{document}
\maketitle
\tableofcontents

\section{Introduction}
\label{sec:introduction}
Recall that a functor between posets is an order-preserving function, and a natural transformation $f \to g$ between two order-preserving functions is a witness for $f \leq g$.
The left Kan extension of $F$ along $K$, if it exists, is the ``smallest'' order-preserving function so that $F \leq \lan_{K}F \circ K$.
\[\input{img/2-01}\]
The universal property demands that for any order-preserving function $G$ such that $F \leq G \circ K$, $\lan_{K}F \leq G$.
This suggests that we take
\[
  \lan_{K}F(d) := \sup\{Fc \mid Kc \leq d\}
\]
Dually, the same argument suggests that we take
\[
  \ran_{K}F(d) := \inf\{Fc \mid d \leq Kc\}
\]
That is, the left (resp., right) Kan extension at $d$ is the supremum (resp., infimum) of all left (resp., right) approximations of $d$ under the image of $F$.
Admittedly, these are the most obvious choices, but they are equivalent to the obvious ones.
Observe that the comma category $K \dn d$ contains pairs $(c, w)$, where $w$ is a witness for $Kc \leq d$.
Thus, the definitions above are equivalent to the followings:
\begin{align}
  \lan_{K}F(d) &:= \colim(\xMor{K \dn d}{\Pi^{d}}{\iC} \xMor{}{F}{\iE})\\
  \ran_{K}F(d) &:= \lim(\xMor{d \dn K}{\Pi_{d}}{\iC} \xMor{}{F}{\iE})
\end{align}

These generalize to any category in which certain limits or colimits exist.

\begin{prop}
  Given functors $F : \iC \to \iE$ and $K : \iC \to \iD$, if for every $d \in \iD$ the colimit
  \[
    \lan_{K}F(d) = \colim(\xMor{K \dn d}{\Pi^{d}}{\iC} \xMor{}{F}{\iE})
  \]
  exists, then they define the left Kan extension.
  Dually, if for every $d \in \iD$ the limit
  \[
    \ran_{K}F(d) = \lim(\xMor{d \dn K}{\Pi_{d}}{\iC} \xMor{}{F}{\iE})
  \]
  exists, then they define the right Kan extension.
\end{prop}
\begin{proof}
  First, we need to extend $\lan_{K}F$ to a functor.
  Consider any morphism $g : d \to d'$.
  This morphism induces a canonical functor $g_{*} : K \dn d \to K \dn d'$ defined by post-composition.
  In particular, the following diagram commutes.
  \[\input{img/2-02}\]
  For each $d \in \iD$, $\lan_{K}F(d)$ provides a specified colimit cocone $\lambda^{d} : F\Pi^{d} \to \lan_{K}F(d)$.
  For $\lambda^{d'} : F\Pi^{d'} \to \lan_{K}F(d')$, we can define a cocone $\mu^{d'} : F\Pi^{d} \to \lan_{K}F(d')$ by reindexing with $g$.
  The universal property then yields a unique morphism $\lan_{K}F(d) \to \lan_{K}F(d')$.
  \[\input{img/2-03}\]
  Uniqueness implies functoriality.
  The unit natural transformation is defined by
  \[
    \eta_{c} := \lambda^{Kc}_{\id_{Kc}} : Fc \to \lan_{K}F(Kc)
  \]
  Consider
  \[\input{img/2-04}\]
  The bottom composite gives $\lambda^{Kc'}_{Kf}$, while by definition $\lan_{K}F(Kf)$ that makes the top triangle commute.
  Thus, $\eta$ is natural.
  It remains to show that $(\lan_{K}F, \eta)$ has the required universal property.
  For any pair $(G, \gamma)$ defining a left extension, we need to factorize $\gamma$.
  Naturality of $\gamma$ allows us to extend it to a cocone under $F\Pi^{d}$.
  \[\input{img/2-05}\]
  Thus, the universal property of colimits yields a unique morphism $\lan_{K}F(d) \to Gd$ so that $\alpha_{d} \circ \lambda^{d}_{f} = Gf \circ \gamma_{c}$.
  We take that unique morphism to be $\alpha_{d}$.
  Clearly,
  \[
    (\alpha \circ \eta)_{c} = \alpha_{Kc} \circ \lambda^{Kc}_{\id_{Kc}} = \gamma_{c}
  \]
\end{proof}

\begin{cor}
  If $K : \iC \to \iD$ is a functor so that $\iC$ is small and $\iD$ is locally small then
  \begin{enumerate}
  \item If $\iE$ is cocomplete, then left Kan extensions $\lan_{K} \adj K^{*}$ exist and are given by the colimit formula.
  \item If $\iE$ is complete, the right Kan extensions $K^{*} \adj \ran_{K}$ exist and are given by the limit formula.
  \end{enumerate}
\end{cor}
\begin{proof}
  Smallness implies that the respective comma categories $K \dn d$ and $d \dn K$ are small.
  Cocompleteness (resp., completeness) of $\iE$ then implies that the necessary colimits (resp., limits) exist.
\end{proof}

\begin{eg}
  Directed graphs are functors from the category with two objects $E, V$ and a parallel pair of morphisms $s, t : E \toto V$ to $\mathsf{Set}$.
  A natural transformations between two such functors is a graph homomorphism.
  The forgetful functor $\mathsf{DirGraph} \to \mathsf{Set}$ that maps a graph to its set of vertices is given by restricting along the functor from the terminal category that picks out the object $V$.
  The forgetful functor admits both left and right adjoints (the free and cofree graphs, respectively).
  These adjoints can be computed as Kan extensions:
  \[\input{img/2-06}\]
  The action on $V$ is trivial since
  \[\mathrm{res} \dn V \iso 1 \iso V \dn \mathrm{res}\]
  Thus, $\lan G(V) = G = \ran G(V)$.
  For edges, $\mathrm{res} \dn E \iso 0$ since there is no morphism $f : V \to E$ and $E \dn \mathrm{res} \iso 2$ since there are exactly two morphisms $s, t : E \toto V$.
  The colimit under an empty diagram is the initial object.
  Thus, $\lan G(E) = \varnothing$.
  This means that $\lan G$ is the discrete graph with vertices $G$.
  
  The colimit under a diagram indexed by the discrete category $2$ is a product.
  Thus, $\ran G(E) = G \times G$.
  This means that $\ran G$ is the complete graph with vertices $G$.
\end{eg}

\bibliographystyle{alpha}
\bibliography{all}

\end{document}
