\documentclass{amsart}
\input{decls}
\title{Groups}
\author{Frank Tsai}
\date{\today}
%\thanks{}
\begin{document}
\maketitle
\tableofcontents

\section{Definition of Group}
\label{sec:definition-of-group}

\begin{defn}
  A \emph{group} consists of the following data:
  \begin{enumerate}
  \item a set $G$;
  \item an \emph{identity element} $e \in G$;
  \item a \emph{group multiplication} $\cdot : G \times G \to G$;
  \item a \emph{group inverse} $(\blank)\inv : G \to G$
  \end{enumerate}
  such that
  \begin{enumerate}
  \item group multiplication is associative;
  \item $a\inv \cdot a = e = a \cdot a\inv$ for all $a \in G$.
  \end{enumerate}
\end{defn}
The group axioms can be stated in terms of commutative diagrams.

\begin{mathpar}
  \input{img/01}\and
  \input{img/02}
\end{mathpar}

\begin{eg}[Symmetry Groups]
\end{eg}

\begin{eg}[Dihedral Groups]
\end{eg}

\section{Order}
\label{sec:order}

\begin{defn}
  An element $g$ of a group $G$ has \emph{finite order} if $g^{n} = e$ for some positive integer $n$.
  In this case, the \emph{order}, denoted $|g|$, of $g$ is the smallest such $n$.
  We write $|g| = \oo$ if $g$ does not have finite order.
  The order $|G|$ of a group $G$ is the cardinality of its underlying set.
\end{defn}

\section{Group Homomorphisms}
\label{sec:group-homomorphisms}

\begin{defn}
  A \emph{group homomorphism} from $(G,m_{G})$ to $(H,m_{H})$ is a function $\varphi : G \to H$ that preserves group multiplication.
\end{defn}
This can be express in term of a commutative diagram in $\mathsf{Set}$:
  \begin{equation}
    \label{eq:group-homomorphism}
    \input{img/03}
  \end{equation}

\begin{lem}
  Let $\varphi : G \to H$ be a group homomorphism, and let $g \in G$ be an element of finite order.
  Then $|\varphi(g)|$ divides $|g|$.
\end{lem}
\begin{proof}
  Suppose that $|g| = n$, then $g^{n} = e$.
  Thus,
  \[
    (\varphi(g))^{n} = \varphi(g^{n}) = \varphi(e) = e
  \]
  Thus, $n$ is a multiple of $|\varphi(g)|$.
\end{proof}

\begin{lem}
  Groups and their homomorphisms assemble into a category $\mathsf{Grp}$.
\end{lem}

\begin{lem}
  Trivial groups are both initial and terminal in $\mathsf{Grp}$.
\end{lem}
\begin{proof}
  It's clear that trivial groups are terminal in $\mathsf{Grp}$ since for any set $G$, the unique function from $G$ in to a singleton set is vacuously a group homomorphism.
  Dually, since group homomorphisms preserve group identity, there's a unique group homomorphism from a singleton set to $G$.
\end{proof}

Thus, trivial groups are \emph{zero objects} of $\mathsf{Grp}$.
This implies that $\mathsf{Grp}(G,H)$ is nonempty for any group $G$ and $H$ since one always has
\input{img/04}
Concretely, this homomorphism maps every element of $G$ to the identity of $H$.

\begin{defn}
  An isomorphism of groups is a group homomorphism $\varphi : G \to H$ whose inverse is also a group homomorphism.
  This means that the underlying function of any group homomorphism is automatically a bijection.
\end{defn}
In fact the converse also holds.

\begin{lem}
  $\varphi : G \to H$ is a group isomorphism if and only if $\varphi$ is a bijection.
\end{lem}
\begin{proof}
  Let $\varphi : G \to H$ be a bijective group homomorphism, we need to show that $\varphi\inv : H \to G$ is a group homomorphism.
  That is, we need to show that the diagram
  \input{img/05}
  commutes.
  Since $\varphi$ is a bijection, it is monic.
  Thus, it suffices to show that $\varphi m_{G} (\varphi\inv \times \varphi\inv) = \varphi \varphi\inv m_{H} = m_{H}$.
  This is immediate.
  \[
    \varphi m_{G}(\varphi\inv \times \varphi\inv) = m_{H}(\varphi \times \varphi) (\varphi\inv \times \varphi\inv) = m_{H}
  \]
\end{proof}

\section{Products}
\label{sec:products}

Since the forgetful functor $U : \mathsf{Grp} \to \mathsf{Set}$ is representable by the group $\dZ$, the underlying sets of products in $\mathsf{Grp}$ are themselves products in $\mathsf{Set}$.
Thus, products in $\mathsf{Grp}$ are formed by giving group structures to products in $\mathsf{Set}$.

\begin{defn}
  Let $G$ and $H$ be groups, the \emph{direct product} $G \times H$ is the group formed by equipping the product of the underlying sets of $G$ and $H$ with the group structure
\begin{align}
  \cdot &: ((g,h),(g',h')) \mapsto (gg',hh')\\
  (\blank)\inv &: (g,h) \mapsto (g\inv,h\inv)
\end{align}
That is, the group operations are defined componentwise.
\end{defn}

\section{Coproducts}
\label{sec:coproducts}

Coproducts in $\mathsf{Grp}$ is more delicate.
In general, there is no reasonable group structure for the coproduct of two sets $G + H$.
Later we will see how to construct the coproduct of two groups using free construction.
The coproduct of two groups $G * H$ is called the \emph{free product} of $G$ and $H$.

\section{Abelian Groups}
\label{sec:abelian-groups}

The category $\mathsf{Ab}$ of abelian groups also has zero objects, but products coincide with coproducts.
The (co)product $G \times H$ of two abelian group is called the \emph{direct sum}, denoted $G \oplus H$, of $G$ and $H$.

\section{Cyclic Groups}
\label{sec:cyclic-groups}

\begin{defn}
  A group $G$ is \emph{cyclic} if it is isomorphic to $\dZ$ or to $\dZ/n\dZ$ for some $n$.
\end{defn}

\section{Free Groups}
\label{sec:free-groups}

Given a set $A$, how do we construct a group that contains $A$ in the most efficient way?
A construction is a set-theoretic function $j : A \to G$ mapping each element of $A$ to a corresponding group element of $G$.
Then, given two constructions $j_{1} : A \to G_{1}$ and $j_{2} : A \to G_{2}$, we can compare them by asking for a group homomorphism such that the diagram
\input{img/06}
commutes.
In this case, the construction $j_{1}$ is more efficient because $j_{2}$ can be recovered from $j_{1}$ via $\varphi$.

The most efficient construction is one such that every construction can be uniquely recovered:
\begin{equation}
  \label{eq:free-group}
  \input{img/07}
\end{equation}

That is,
\[
  \mathsf{Grp}(F(A),G) \iso \mathsf{Set}(A,U(G))
\]
Concretely, the underlying set of $F(A)$ is the quotient of all formal words whose letters are $a$ and $a\inv$ for each $a \in A$.
Words are quotient by the congruence that $a\inv a = \epsilon = a a\inv$ for all $a \in A$.

Let $[w]$ and $[w']$ be elements of $F(A)$, group multiplication is defined as $[w \cdot w']$, i.e., the concatenation of $w$ and $w'$ modulo the congruence.
The inverse replaces each letter in a word with its inverse.

The construction $\eta$ maps each element $a \in A$ to the singleton word $[a] \in F(A)$.
\eqref{eq:free-group} forces us to define $\varphi([a]) = f(a)$ for any $a \in A$, while \eqref{eq:group-homomorphism} forces us to define $\varphi([w]) = f(w_{1}) \cdot \cdots \cdot f(w_{n})$.


\end{document}
